# GAM with interaction terms

There are two ways to include interactions between variables:

- For two smoothed variables, the syntax would be: `s(x1, x2)`
- For one smoothed variable and one linear variable (either factor or continuous), the syntax would use the `by` argument `s(x1, by = x2)`:
  - When `x2` is a factor, you have a smooth term that vary between different levels of `x2`;
  - When `x2` is continuous, the linear effect of `x2` varies smoothly with `x1`;
  - When `x2` is a factor, the factor needs to be added as a main effect in the model.


## Interaction between smoothed and factor variables

We will examine interaction effects to determine whether the non-linear smoother `s(SampleDepth)` varies across different levels of `Season`.

```{r}
factor_interact <- gam(Sources ~ Season + 
                         s(SampleDepth, by=Season) + 
                         s(RelativeDepth), 
                       data = isit, method = "REML")

summary(factor_interact)$s.table
```

```{r, fig.height = 8}
par(mfrow = c(2,2))
plot(factor_interact)
```

:::explanation
The first two panels show the interaction effect of the `SampleDepth` smooth and each level of our factor variable, `Season`. Do you see a difference between the two smooth curves?
:::

The plots show some differences between the shape of the smooth terms among the two levels of `Season`. The most notable difference is the peak in the second panel, which tells us that there is an effect of `SampleDepth` between 1000 and 2000 that is important in Season 2, but does not occur in Season 1. This hints that the interaction effect could be important to include in our model.

We can also plot the interaction effect in 3D on a single plot, using `vis.gam()`.

```{r}
vis.gam(factor_interact, theta = 120, n.grid = 50, lwd = .4)
```
> This plot can be rotated by changing the value of the `theta` argument.


To test our idea that this interaction is important, we will perform a model comparison using AIC to determine whether the interaction term improves our model's performance.

```{r}
AIC(two_smooth_model, factor_interact)
```

The AIC of our model with a factor interaction between the `SampleDepth` smooth and `Season` has a lower AIC score, which tells us this model performs better than `two_smooth_model`. Including this interaction seems to improve our model's performance.


## Interaction between smoothed variables

Next, we'll look at the interactions between two smoothed terms, `SampleDepth` and `RelativeDepth`.

```{r}
smooth_interact <- gam(Sources ~ Season + s(SampleDepth, RelativeDepth), 
                       data = isit, method = "REML")
summary(smooth_interact)$s.table
```

In the previous section, we were able to visualise an interaction effect between a smooth and a factor term by plotting a different smooth function of `SampleDepth` for each level of `Season`. 

In this model, we have two smoothed terms, which means that the effect of `SampleDepth` varies smoothly with `RelativeDepth`, and vice-versa. When we visualise this interaction, we instead get a gradient between two continuous smooth functions:

```{r}
plot(smooth_interact, page = 1, scheme = 2)
```

We can also plot this interaction on a 3D surface:

```{r}
vis.gam(smooth_interact, 
        view = c("SampleDepth", "RelativeDepth"), 
        theta = 50, n.grid = 50, lwd = .4)
```
> Remember, this plot can be rotated by changing the value of the `theta` argument.

:::explanation
You can change the colour of the 3D plot using the `color` argument. Try specifying `color = "cm"` in `vis.gam()` above, and check `?vis.gam` for more color options.
:::

The plots do illustrate a non-linear interaction effect, where `Sources` is lowest at high values of `SampleDepth` and `RelativeDepth`, but increases with `RelativeDepth` while `SampleDepth` is low.

So, there does seem to be an interaction effect between these smooth terms. Does including the interaction between `s(SampleDepth)` and `s(RelativeDepth)` improve our `two_smooth_model` model?

```{r}
AIC(two_smooth_model, smooth_interact)
```

The model with the interaction between `s(SampleDepth)` and `s(RelativeDepth)` has a lower AIC, which means including this interaction improves our model's performance, and our ability to understand the drivers of bioluminescence. 