# Quick intro to Generalized Additive Mixed Models (GAMMs)

When observations are not independent, GAMs can be used to either incorporate:

-   a serial correlation structure to model residual autocorrelation
    (autoregressive: AR; moving average: MA; or a combination of the two:
    ARMA),
-   random effects that model independence among observations from the
    same site.

That is, in addition to changing the basis as with the `nottem` example, we can also add complexity to the model by incorporating an autocorrelation structure or mixed effects using the `gamm()` function in the `mgcv` package. Although we will not be using it here, the [`gamm4`](https://cran.r-project.org/web/packages/gamm4/gamm4.pdf) package can also be used to estimate GAMMs in R. 

## Residual autocorrelation

To start, let's have a look at a model with temporal autocorrelation in the residuals. We will revisit the Nottingham temperature model and test for correlated errors using the (partial) autocorrelation function.

```{r, echo = TRUE, eval = FALSE}
par(mfrow = c(1,2))
acf(resid(year_gam), lag.max = 36, main = "ACF")
pacf(resid(year_gam), lag.max = 36, main = "pACF")
```

The __autocorrelation function__ (ACF; first panel above) evaluates the cross correlation of a time series with itself at different time lags (i.e. similarity between observations at increasingly large time lags). 

In contrast, the __partial autocorrelation function__ (PACF: second panel above)
gives the partial correlation of a time series with its own lagged values,
after controlling for the values of the time series at all shorter lags.

The ACF and pACF plots are thus used to identify the time steps needed before observations are independent from one another (i.e., when they are no longer autocorrelated).

The ACF plot of our model residuals suggests a significant lag of 1, and perhaps
a lag of 2. Therefore, a low-order AR model is likely needed. 

We can test this by adding AR structures to the Nottingham temperature model. Let's make one with an `AR(1)` (correlation at 1 time step) and one with an `AR(2)` (correlation at two times steps), and compare models using AIC to find a the best-fit model.

```{r}
df <- data.frame(nottem, nottem_year, nottem_month)

year_gam <- gamm(nottem ~ s(nottem_year) + s(nottem_month, bs = "cc"), data = df)

year_gam_AR1 <- gamm(nottem ~ s(nottem_year) + s(nottem_month, bs = "cc"),
                     correlation = corARMA(form = ~ 1|nottem_year, p = 1),
                     data = df)

year_gam_AR2 <- gamm(nottem ~ s(nottem_year) + s(nottem_month, bs = "cc"),
                     correlation = corARMA(form = ~ 1|nottem_year, p = 2),
                     data = df)
```

Which of these models performs the best?

```{r}
AIC(year_gam$lme, year_gam_AR1$lme, year_gam_AR2$lme)
```

The `AR(1)` provides a significant increase in fit over the naive model (`year_gam`), but there is very little improvement in moving to the `AR(2)`. So, it is best to include only the `AR(1)` structure in our model.

## Random effects

As we saw in the section about changing the basis, `bs` specifies the type of underlying base function. For random intercepts and linear random slopes we use `bs = "re"`, but for random smooths we use `bs = "fs"`.

There are **three different types of random effects** in GAMMs. Below, we use `fac` to indicate factor coding for the random effect, and `x0` for a continuous fixed effect:

- **Random intercepts** adjust the height of other model terms with a constant value: `s(fac, bs = "re")`
- **Random slopes** adjust the slope of the trend of a numeric predictor: `s(fac, x0, bs = "re")`
- **Random smooths** adjust the trend of a numeric predictor in a nonlinear way: `s(x0, fac, bs = "fs", m = 1)`, where the argument `m = 1` sets a heavier penalty for the smooth moving away from 0, causing shrinkage to the mean.

:::explanation
For more detailed explanations about random effects, see [Workshop 7](https://r.qcbs.ca/workshops/r-workshop-07/).
:::

:::noway
This is a (very!) brief overview of how you can use random effects in GAMMs. For further details, we _really_ recommend @pedersen2019hierarchical, which is a very accessible article outlining various ways to build GAMMs to answer ecological questions.
:::

### GAMM with a random intercept

We will use the `gamSim()` function to generate a dataset with a random effect, then run a model with a random intercept using `fac` as the random factor.

```{r}
# generate and view data
gam_data2 <- gamSim(eg = 6)
head(gam_data2)
```

```{r}
# run random intercept model
gamm_intercept <- gam(y ~ s(x0) + s(fac, bs = "re"), data = gam_data2, method = "REML")

# examine model output
summary(gamm_intercept)$s.table
```

Note that there is now a smoother term for the random intercept in the
summary table. You can plot and view the random intercepts for each
level of `fac` as follows:

```{r}
plot(gamm_intercept, select = 2) 
# select = 2 because the random effect appears as the second entry in the summary table.
```

We can also use the `plot_smooth` function to visualize the model, which allows us to plot a smooth of the
summed effects of a GAM (based on predictions). This function also optionally removes the random effects by setting `rm.ranef = TRUE`. 

Here, we will plot the summed effects for the `x0` without random effects, and then plot the predictions of all four levels of the random `fac` effect:

```{r, results='hide', fig.width=10, fig.height=6}
par(mfrow = c(1,2), cex = 1.1)

# Plot the summed effect of x0 (without random effects)
plot_smooth(gamm_intercept, view = "x0", rm.ranef = TRUE, 
            main = "intercept + s(x1)")

# Plot each level of the random effect
plot_smooth(gamm_intercept, view = "x0", rm.ranef = FALSE,
            cond = list(fac="1"),
            main = "... + s(fac)", col = 'orange', ylim = c(0,25))
plot_smooth(gamm_intercept, view = "x0", rm.ranef = FALSE,
            cond = list(fac = "2"), 
            add = TRUE, col = 'red')
plot_smooth(gamm_intercept, view="x0", rm.ranef = FALSE,
            cond = list(fac = "3"), 
            add = TRUE, col = 'purple')
plot_smooth(gamm_intercept, view="x0", rm.ranef = FALSE,
            cond = list(fac = "4"), 
            add = TRUE, col = 'turquoise')
```


### GAMM with a random slope

Next, we will run and plot a model with a random slope:

```{r}
gamm_slope <- gam(y ~ s(x0) + s(x0, fac, bs = "re"), data = gam_data2, method = "REML")

summary(gamm_slope)$s.table
```

We can once again plot the summed effects for the `x0` without random effects, and then plot the predictions of all four levels of the random `fac` effect:

```{r, results='hide', fig.width=10, fig.height=6}
par(mfrow = c(1,2), cex = 1.1)

# Plot the summed effect of x0 (without random effects)
plot_smooth(gamm_slope, view = "x0", rm.ranef = TRUE, 
            main = "intercept + s(x1)")

# Plot each level of the random effect
plot_smooth(gamm_slope, view = "x0", rm.ranef = FALSE,
            cond = list(fac="1"),
            main = "... + s(fac, x0)", col = 'orange', ylim = c(0,25))
plot_smooth(gamm_slope, view = "x0", rm.ranef = FALSE,
            cond = list(fac = "2"), 
            add = TRUE, col = 'red')
plot_smooth(gamm_slope, view="x0", rm.ranef = FALSE,
            cond = list(fac = "3"), 
            add = TRUE, col = 'purple')
plot_smooth(gamm_slope, view="x0", rm.ranef = FALSE,
            cond = list(fac = "4"), 
            add = TRUE, col = 'turquoise')
```


### GAMM with a random intercept and slope

We will now include both a random intercept _and_ a random slope term.

```{r}
gamm_int_slope <- gam(y ~ s(x0) + s(fac, bs = "re") + s(fac, x0, bs = "re"),
                      data = gam_data2, method = "REML")

summary(gamm_int_slope)$s.table
```

We can once again plot the summed effects for the `x0` without random effects, and then plot the predictions of all four levels of the random `fac` effect:

```{r, results='hide', fig.width=10, fig.height=6}
par(mfrow = c(1,2), cex = 1.1)

# Plot the summed effect of x0 (without random effects)
plot_smooth(gamm_int_slope, view = "x0", rm.ranef = TRUE, 
            main = "intercept + s(x1)")

# Plot each level of the random effect
plot_smooth(gamm_int_slope, view = "x0", rm.ranef = FALSE,
            cond = list(fac="1"),
            main = "... + s(fac) + s(fac, x0)", col = 'orange', ylim = c(0,25))
plot_smooth(gamm_int_slope, view = "x0", rm.ranef = FALSE,
            cond = list(fac = "2"), 
            add = TRUE, col = 'red')
plot_smooth(gamm_int_slope, view="x0", rm.ranef = FALSE,
            cond = list(fac = "3"), 
            add = TRUE, col = 'purple')
plot_smooth(gamm_int_slope, view="x0", rm.ranef = FALSE,
            cond = list(fac = "4"), 
            add = TRUE, col = 'turquoise')
```

Note that the random slope is static in this case:

```{r}
plot(gamm_int_slope, select=3) 
# select = 3 because the random slope appears as the third entry in your summary table.
```

### GAMM with a random smooth

Lastly, we will examine a model with a random smooth.

```{r}
gamm_smooth <- gam(y ~ s(x0) + s(x0, fac, bs = "fs", m = 1), 
                   data = gam_data2, method = "REML")

summary(gamm_smooth)$s.table
```

Here, if the random slope varied along `x0`, we would see different curves for each level:

```{r, echo = TRUE, eval = FALSE}
plot(gamm_smooth, select=1) 
# select = 1 because the smooth slope appears as the first entry in your summary table.
```

We can once again plot the summed effects for the `x0` without random effects, and then plot the predictions of all four levels of the random `fac` effect:

```{r, results='hide', fig.width=10, fig.height=6}
par(mfrow = c(1,2), cex = 1.1)

# Plot the summed effect of x0 (without random effects)
plot_smooth(gamm_smooth, view = "x0", rm.ranef = TRUE, 
            main = "intercept + s(x1)")

# Plot each level of the random effect
plot_smooth(gamm_smooth, view = "x0", rm.ranef = FALSE,
            cond = list(fac="1"),
            main = "... + s(fac) + s(fac, x0)", col = 'orange', ylim = c(0,25))
plot_smooth(gamm_smooth, view = "x0", rm.ranef = FALSE,
            cond = list(fac = "2"), 
            add = TRUE, col = 'red')
plot_smooth(gamm_smooth, view="x0", rm.ranef = FALSE,
            cond = list(fac = "3"), 
            add = TRUE, col = 'purple')
plot_smooth(gamm_smooth, view="x0", rm.ranef = FALSE,
            cond = list(fac = "4"), 
            add = TRUE, col = 'turquoise')
```


### GAMM model comparison

All of the mixed models from this section can be compared using `AIC()` to determine the best fit model

```{r}
AIC(gamm_intercept, gamm_slope, gamm_int_slope, gamm_smooth)
```

The best model among those we have built here would be a GAMM with a random effect on the intercept.